---
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
[![Last-changedate](https://img.shields.io/badge/last%20change-`r gsub('-', '--', Sys.Date())`-brightgreen.svg)](https://github.com/pat-s/pathogen-modeling/commits/master) [![minimal R version](https://img.shields.io/badge/R%3E%3D-3.5.0-brightgreen.svg)](https://cran.r-project.org/)   [![Licence](https://img.shields.io/github/license/mashape/apistatus.svg)](http://choosealicense.com/licenses/mit/)  

## Reproducible analysis for paper:

##### tailfindr: Alignment-free poly(A) length measurement for Oxford Nanopore RNA and DNA sequencing

Maximilian Krause,  Adnan M. Niazi,  Kornel Labun,  Yamila N. Torres Cleuren,  Florian S. Müller,  Eivind Valen

## About the repo

This repository is organized as an R package, providing functions and the raw data to reproduce and extend the analysis reported in the publication. By raw data, we mean the output of tools such as tailfindr and Nanopolish etc.

This project is setup with a drake workflow, ensuring reproducibility. Intermediate targets/objects will be stored in a hidden .drake directory.

The R library of this project is managed by packrat. This makes sure that the exact same package versions are used when recreating the project. When calling packrat::restore(), all required packages will be installed with their specific version.

Please note that this project was built with R version 3.6.0 on a MAC OSx Mojave operating system. The packrat packages from this project are not compatible with R versions prior version 3.6.0 (In general, it should be possible to reproduce the analysis on any other operating system.)

## Getting started

To clone the project, a working installation of git is required. Open a terminal in the directory of your choice and execute:

```sh
git clone https://github.com/adnaniazi/krauseNiazi2019Analyses.git
```
Then go into the `krauseNiazi2019Analyses` directory using:
```sh
cd krauseNiazi2019Analyses
```
Now start R in this directory and run:

```{r README-4, eval = FALSE}
packrat::restore() # restores all R packages with their specific version
drake::r_make()    # recreates the analysis
```
You can also run the above two commands in RStudio by first opening krauseNiazi2019Analyses folder as project and executing the two commands in R console.

The first command restores all the packages that we have used for the analyses. The second command builds all the analyses. It first downloads all the data in CSV from into the `extdata` directory, then it consolidates information from different CSV into a single data frame. This consolidated data frame is then used in the Rmarkdown files to generate the plots.